Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-05-03 doc/invt.tex
invt.tex: editing gamcmp.gp and introduction

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the task is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\ln p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Normally, the distribution $p^*(z)$ is
limited in one or a few narrow $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the entire curve.
%
This solution can be laborious % clumsy in practice
and it often fails to converge quickly
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum.
%for a long time.

A better solution is to artificially flatten
the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%
This alleviates the above problem of local trapping
by forcing the system to travel more
deliberately along $z$,
which often represents
a direction of slow motion.
%
To achieve a flat distribution, however,
the method must be able to construct a
bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is gradually built up
during simulation.
%
These techniques employ updating loops
to discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
The main difference
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian window
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
%which % brings about a side effect of disrupting
and the fluctuation
disrupts \note{``disturbs''?} the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the effect.

The updating schedule,
or the manner of reducing
the updating magnitude,
therefore,
determines the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
(with ``time'' being
the number of simulation steps).
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive flat-distribution sampling method.
%
In this method, the optimal schedule is derived
from an equation of motion of a free particle,
with the initial updating magnitude set to
half of the previous equilibrium value.
%
The optimal schedule in the WL case
is shown to be given by
the inverse-time formula,
with the origin of time modified
to satisfy the above optimal initial condition.
%
For metadynamics, however,
the optimal schedule is more complex,
and it can be sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the modified inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of
flat-distribution sampling
and some known aspects on the optimal schedule.
%
This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin (WL) updating scheme.
%
We shall prove the optimality
of the known inverse-time formula.
%
Next, we turn to the general case
of multiple-bin updating schemes
and propose a method of computing
the optimal schedule
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

@@

To demonstrate the optimal schedule from Eq. \eqref{eq:q_opt},
we used the Gaussian (with $\sigma = 10$) updating scheme
as an example.
%
In Table \ref{tab:error_Gaussian},
we showed that the optimal schedule
produced smaller errors than
the inverse-time schedules.
%
However, the optimal schedules
decayed more slowly than
the inverse-time formula, Eq. \eqref{eq:alpha_invt1},
and depended on the sampling process,
as shown in Fig. \ref{fig:optacmp}.
%
The impeded decay of the optimal schedule
was necessary in order to
damp out the short-wavelength eigenmodes
(i.e., local noises),
as shown in Fig. \ref{fig:xerr}.
%
If, however, we could ignore these local eigenmodes
(by explicitly filtering out
them in the biased potential),
then we can modify the optimal schedule
from Eq. \eqref{eq:q_opt}
and limit the sum
to the first few eigenmodes.
%
The resulting schedule
would became similar to the inverse-time schedule,
Eq. \eqref{eq:alpha_invt1},
as shown in Fig. \ref{fig:optacmp}.
%
\note{A more thorough approach is
to use the bandpass updating scheme
to be discussed later.
}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $K \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the same shape,
    but meet the initial error curve
    at a lower level (not shown).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation,
e.g., the WL algorithm or metadynamics.
%
Adaptive FDS methods
effectively sample a flat distribution
by incrementally updating the bias potential
to offset the PMF.
%
The optimal schedule delivers fastest convergence
of the bias potential.
%They differ by the updating schemes,
%which specify how the bias potential is modified
%in each updating step.
%%
%In the single-bin or WL case,
%the update is limited to the current bin;
%while in the metadynamics case,
%the bias potential of a neighborhood of the current one
%is updated with relative weight specified by a Gaussian window.

@@

Strictly speaking, our method is valid
only for long simulations.
%
This allows us to assume the sampling process
to be a quasi-equilibrium one,
and to represent the autocorrelation function
of each eigenmode by a single number, $\Gamma_k$.
%
Although these assumptions appeared to work well
on our test system,
they may break down for relatively short simulations
of glassy systems.

@@

As an example, we used the method to compute
the integrals, $\Gamma_k$'s, from an MD simulation,
and compared them to those from
%Eq. \eqref{eq:Gamma_onestep}.
the local MC sampling process.
%
As shown in Fig. \ref{fig:gamcmp},
the larger values of $\Gamma_k$'s
with smaller $k$'s
from the two sampling processes
were almost proportional.
%
This suggests that
the optimal schedule
from simulations using local sampling processes
would be similar in general properties.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the local MC simulation.
    %
    Each simulation was ran for $10^{10}$ steps.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$.
    %
    For the MD simulation,
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The velocity rescaling thermostat\cite{bussi2007} was
    applied for the unit reduced temperature
    every MD step with viscosity $20$.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      The data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

The error of the equilibrium FDS simulation
depends on the precision of the histogram.
%
If all $v_i$'s are exact,
the error is minimized, and can be computed
from the second term
on the right-hand side of Eq. \eqref{eq:vcorr_equil},
%
\begin{align}
  E
  &=
  \left\langle
    \sum_{ i = 1 }^ n
      \rho_i \,
      \left(
        \ln \frac { H_i }
                  { p_i }
        -
        \sum_{r = 1}^n \rho_r
        \ln \frac { H_r }
                  { p_r }
      \right)^2
  \right\rangle
  \notag
  \\
  &
  \approx
  \sum_{ i = 1 }^ n
    \rho_i \,
    \left\langle
      \left(
        \frac { H_i }
              { p_i }
        -
        \sum_{r = 1}^n \rho_r
            \frac { H_r }
                  { p_r }
      \right)^2
    \right\rangle,
\notag
\end{align}
where
$H_i = \frac{1}{T} \sum_{t = 1}^T h_i(t)$
with
$h_i(t)$ defined by Eq. \eqref{eq:h_def}.

>>>>

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the task is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a collective variable $z$.
%
However, $z$
often represents a slow variable
with a localized or multi-modal distribution, $p^*(z)$.
%
Thus,
to capture the global shape of the free energy surface
in terms of the potential of mean force (PMF), $-\ln p^*(z)$,
an effective strategy is to sample a flat distribution
along $z$ by introducing a bias potential that
cancels the PMF.
%The negative logarithm,
%$-\ln p^*(z)$,
%defines a free energy,
%or the potential of mean force (PMF).
%%
%Normally, the distribution $p^*(z)$ is
%limited in one or a few narrow $z$ regions.
%%
%Therefore, to capture the global shape
%of the PMF,
%one has to run multiple independent simulations
%under different umbrella or bias potentials
%along $z$\cite{mezei1987, berg1992, lee1993},
%and then piece together the entire curve.
%%
%This solution can be laborious % clumsy in practice
%and it often fails to converge quickly
%for a complex and glassy system,
%because the system can be trapped
%in a local free energy minimum.
%%for a long time.
%
%
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
%wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential on the fly.
%
These techniques employ regular updating loops
to discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
%The main difference
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
However,
since the frequent updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying
equilibrium sampling dynamics\cite{laio2002},
one often has to gradually decrease
the updating magnitude
to reduce the effect.

The schedule of reducing
the updating magnitude,
therefore,
determines the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
(with ``time'' being
the number of simulation steps).
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive FDS method,
including the WL algorithm and metadynamics.
%
We derive the optimal schedule
from an equation of motion of a free particle,
and show that the initial updating magnitude
should be set to
half of the previous equilibrium value.
%
The optimal schedule in the WL case
is shown to be given by
the inverse-time formula,
with a modified origin of time.
%
However,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the modified inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help practitioners
plan and optimize FDS simulations.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects on the optimal schedule.
%
%This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin (WL) updating scheme.
%
We shall prove the optimality
of the known inverse-time formula.
%
Next, we turn to the general case
of multiple-bin updating schemes
and propose a method of computing
the optimal schedule
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$,
an FDS simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

@@

To demonstrate the optimal schedule from Eq. \eqref{eq:q_opt},
we used the Gaussian (with $\sigma = 10$) updating scheme
as an example.
%
In Table \ref{tab:error_Gaussian},
we showed that the optimal schedule
produced smaller errors than
the inverse-time schedules.
%
However, the optimal schedules
decayed more slowly than
the inverse-time formula, Eq. \eqref{eq:alpha_invt1},
and depended on the sampling process,
as shown in Fig. \ref{fig:optacmp}.
%
The impeded decay of the optimal schedule
was necessary in order to
damp out the short-wavelength eigenmodes
(i.e., local noises),
as shown in Fig. \ref{fig:xerr}.
%
If, however, we could ignore these local eigenmodes
(by explicitly filtering
them from the biased potential),
then we can modify the optimal schedule
from Eq. \eqref{eq:q_opt}
and limit the sum
to the first few eigenmodes.
%
The resulting schedule
would became similar to the inverse-time schedule,
Eq. \eqref{eq:alpha_invt1},
as shown in Fig. \ref{fig:optacmp}.
%
\note{A more thorough approach is
to use the bandpass updating scheme
to be discussed later.
}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Predicted error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $K \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the same shape,
    but meet the initial error curve
    at a lower level (not shown).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation,
e.g., the WL algorithm or metadynamics.
%
Adaptive FDS methods
effectively sample a flat distribution
by incrementally updating the bias potential
to offset the PMF.
%
The optimal schedule delivers the fastest convergence
of the bias potential,
and thus helps minimize the computational effort
of the adaptive FDS simulation.
%They differ by the updating schemes,
%which specify how the bias potential is modified
%in each updating step.
%%
%In the single-bin or WL case,
%the update is limited to the current bin;
%while in the metadynamics case,
%the bias potential of a neighborhood of the current one
%is updated with relative weight specified by a Gaussian window.

@@

Strictly speaking, our method is valid
only for long simulations.
%
This allows us to assume the sampling process
to be a quasi-equilibrium one\cite{barducci2008, dama2014},
and to represent the autocorrelation function
of each eigenmode by a single number, $\Gamma_k$.
%
Although these assumptions appeared to work well
on our test system,
they may break down for relatively short simulations
of glassy systems.

@@

As an example, we used the method to compute
the integrals, $\Gamma_k$'s, from
a local MC simulation
(cf. Sec. \ref{sec:results_system})
and an MD simulation,
and compared them with those from
the one-step sampling process.
%
As shown in Fig. \ref{fig:gamcmp},
the $\Gamma_k$'s of the small-$k$ eigenmodes
from the three sampling processes
were almost proportional,
suggesting that
the optimal schedule
for simulations using local sampling processes
would be similar in general properties.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from a local MC simulation
    and an MD simulation.
    %
    Each simulation was ran for $10^{10}$ steps.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$.
    %
    For the MD simulation,
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The velocity rescaling thermostat\cite{bussi2007} was
    applied for the unit reduced temperature
    every MD step with viscosity $20$.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      The data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

The error of the equilibrium FDS simulation
depends on the precision of the histogram.
%
If all $v_i$'s are exact,
the error is minimized, and can be computed
from the second term
on the right-hand side of Eq. \eqref{eq:vcorr_equil} as
%
\begin{align}
  E
  &=
  \left\langle
    \sum_{ i = 1 }^ n
      \rho_i \,
      \left(
        \ln \frac { H_i }
                  { p_i }
        -
        \sum_{r = 1}^n \rho_r
        \ln \frac { H_r }
                  { p_r }
      \right)^2
  \right\rangle
  \notag
  \\
  &
  \approx
  \sum_{ i = 1 }^ n
    \rho_i \,
    \left\langle
      \left(
        \frac { H_i }
              { p_i }
        -
        \sum_{r = 1}^n \rho_r
            \frac { H_r }
                  { p_r }
      \right)^2
    \right\rangle,
\notag
\end{align}
where
$H_i = \frac{1}{T} \sum_{t = 1}^T h_i(t)$
with
$h_i(t)$ defined by Eq. \eqref{eq:h_def}.
